---
title: "Headspace GC-MS Data Analaysis Chardonnay Fermentations"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This analysis is for the GCMS headspace data collected at Supra. The peak areas are from confirmed compounds with standards ran as well as some putitativly identified peaks. 

The peak areas were normalized to the TIC for each sample, therefore the values being analyzed in this dataset are RELATIVE ABUNDANCES and NOT absolute quantitative values. 


```{r message=FALSE}
library(readxl)
library(tigerstats)
library(tidyverse)
library(ggplot2)
library(ggpubr)
library(visreg)
library(car)
library(agricolae)
library(devtools)
library(xlsx)
library(corrplot)
library(Hmisc)
library(cowplot)
library(ggcorrplot)
library(RColorBrewer)
library(gplots)
library(ggfortify)
library(ggthemes)
```


```{r}
df<-read_xlsx("revised peak areas 20200314.xlsx", sheet= "TIC_Norm_areas") %>% data.frame()
df <- filter(df, strain != "curve")
df <- filter(df, strain != "Model_Wine")
df <- filter(df, strain != "Blank")
df <- filter(df, strain != "QC")

full <- df
hot <- filter(df, temperature == "25")
cold <- filter(df, temperature =="15")
```




______________________________________________________________________________________________________________________________

Create PCA plots for the full data set colored by strain and temperature. Blanks, QCs were removed. the plots are saved as JPEGS, and loading plots are omitted.  
```{r}
# PCA
full_pcaformat <- full[6:ncol(full)] %>% data.frame()
row.names(full_pcaformat) <- full$sample_ID
full_pca <- prcomp(full_pcaformat, center = TRUE, scale. = TRUE)


#Colored by strain
strain <- full$strain
all_species_pca <- autoplot(full_pca, data = full, colour = 'species', scale = T, size = 1.25,
         loadings = FALSE, loadings.label = FALSE,loadings.label.repel=T,
         loadings.label.size = 3,loadings.colour = 'black', 
         loadings.label.colour = 'black') +
  theme(panel.background = element_blank(),
        panel.border=element_rect(fill=NA),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        strip.background=element_blank(),
        axis.text.x=element_text(colour="black"),
        axis.text.y=element_text(colour="black"),
        axis.ticks=element_line(colour="black"),
        plot.margin=unit(c(1,1,1,1),"line"),
        legend.key.size = unit(.025, "cm"),
        legend.key.width = unit(.25,"cm"))+
  scale_color_colorblind(labels = c(expression(italic("S. cerevisiae")), expression(italic("S. uvarum")))) 
 
#ggsave("output/pca_plots/pca_full_Strainv2.jpeg")

#Colored by Temperature
#temperature <- full$temperature
all_temp_pca <- autoplot(full_pca, data = full, colour = 'temperature', scale = T, size = 1.25,
         loadings = FALSE, loadings.label = FALSE,loadings.label.repel=T,
         loadings.label.size = 3,loadings.colour = 'black', 
         loadings.label.colour = 'black') +
  theme(panel.background = element_blank(),
        panel.border=element_rect(fill=NA),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        strip.background=element_blank(),
        axis.text.x=element_text(colour="black"),
        axis.text.y=element_text(colour="black"),
        axis.ticks=element_line(colour="black"),
        plot.margin=unit(c(1,1,1,1),"line"),
        legend.key.size = unit(.025, "cm"),
        legend.key.width = unit(.25,"cm")) +
  scale_color_colorblind(labels = c(expression("15C"), expression("25C"))) 
#ggsave("pca_full_temp.jpeg")
```



______________________________________

PCA plots of 15C (cold) dataset:
```{r}

cold_pcaformat <- cold[6:ncol(cold)] %>% data.frame()
row.names(cold_pcaformat) <- cold$sample_ID
cold_pca <- prcomp(cold_pcaformat, center = TRUE, scale. = TRUE)
#summary(cold_pca)
cold$strain
#PCA plot of cold temp fermentations colored by strain:
strain <- cold$strain
cold_strain_pca <- autoplot(cold_pca, data = cold, colour = 'strain', scale = TRUE, size = 1.25,
         loadings = FALSE, loadings.label = FALSE,loadings.label.repel=T,
         loadings.label.size = 3,loadings.colour = 'black', 
         loadings.label.colour = 'black') +
  theme(panel.background = element_blank(),
        panel.border=element_rect(fill=NA),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        strip.background=element_blank(),
        axis.text.x=element_text(colour="black"),
        axis.text.y=element_text(colour="black"),
        axis.ticks=element_line(colour="black"),
        plot.margin=unit(c(1,1,1,1),"line"),
        legend.key.size = unit(.025, "cm"),
        legend.key.width = unit(.25,"cm"),
        legend.text=element_text(size=6.5))+ scale_color_calc()
 


#ggsave("output/pca_plots/pca_15C_strain.jpeg")
 
#make a PCA plot of the cold dataset colored by cerevisiae or uvarum:
species <- cold$species
cold_species_pca <- autoplot(cold_pca, data = cold, colour = 'species', scale = TRUE, size = 1.25,
         loadings = FALSE, loadings.label = FALSE,loadings.label.repel=T,
         loadings.label.size = 3,loadings.colour = 'black', 
         loadings.label.colour = 'black') +
  theme(panel.background = element_blank(),
        panel.border=element_rect(fill=NA),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        strip.background=element_blank(),
        axis.text.x=element_text(colour="black"),
        axis.text.y=element_text(colour="black"),
        axis.ticks=element_line(colour="black"),
        plot.margin=unit(c(1,1,1,1),"line"),
        legend.key.size = unit(.025, "cm"),
        legend.key.width = unit(.25,"cm")) +
  scale_color_colorblind(labels = c(expression(italic("S. cerevisiae")), expression(italic("S. uvarum")))) 

#ggsave("output/pca_plots/pca_15C_species.jpeg")
```
labels = c(expression(italic("S. cerevisiae")), expression(italic("S. uvarum")))

___________________________________

PCA plots of 25C (hot) dataset:
```{r}
hot_pcaformat <- hot[6:ncol(hot)] %>% data.frame()
row.names(hot_pcaformat) <- hot$sample_ID

hot_pca <- prcomp(hot_pcaformat, center = TRUE, scale. = TRUE)
#summary(hot_pca)

#Hot df colored by strain
strain <- hot$strain
hot_strain_pca <- autoplot(hot_pca, data = hot, colour = 'strain', scale = TRUE, size = 1.25,
         loadings = FALSE, loadings.label = FALSE,loadings.label.repel=T,
         loadings.label.size = 3,loadings.colour = 'black', 
         loadings.label.colour = 'black') +
  theme(panel.background = element_blank(),
        panel.border=element_rect(fill=NA),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        strip.background=element_blank(),
        axis.text.x=element_text(colour="black"),
        axis.text.y=element_text(colour="black"),
        axis.ticks=element_line(colour="black"),
        plot.margin=unit(c(1,1,1,1),"line"),
        legend.key.size = unit(.025, "cm"),
        legend.key.width = unit(.25,"cm"),
        legend.text=element_text(size=6.5)) + scale_color_calc()
#ggsave("output/pca_plots/pca_25C_strain.jpeg")

#make PCA plot of hot fermentations colored by strain type
species <- hot$type
hot_species_pca <-autoplot(hot_pca, data = hot, colour = 'species', scale = TRUE, size = 1.25,
         loadings = FALSE, loadings.label = FALSE,loadings.label.repel=T,
         loadings.label.size = 3,loadings.colour = 'black', 
         loadings.label.colour = 'black') +
  theme(panel.background = element_blank(),
        panel.border=element_rect(fill=NA),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        strip.background=element_blank(),
        axis.text.x=element_text(colour="black"),
        axis.text.y=element_text(colour="black"),
        axis.ticks=element_line(colour="black"),
        plot.margin=unit(c(1,1,1,1),"line"),
        legend.key.size = unit(.025, "cm"),
        legend.key.width = unit(.25,"cm")) +
  scale_color_colorblind(labels = c(expression(italic("S. cerevisiae")), expression(italic("S. uvarum")))) 
#ggsave("output/pca_plots/pca_25C_plot_species.jpeg")
```

Arrange figures into a grid and made into a single figure.
```{r}
pca_grid<-plot_grid(all_temp_pca, all_species_pca,
                    cold_strain_pca, cold_species_pca, 
                    hot_strain_pca, hot_species_pca,
                    ncol = 2, 
                    labels =c('A', 'B', 'C', 'D', 'E', 'F'),
                    label_size = 10,
                    align = "hv"
                    )
pca_grid

#ggsave("output/pca_plots/pca_gridV4.jpeg")
```



Analyze the differences between each strain WITHIN a temperature using an ANOVA. 

- Test for homogeneity of variance - levene's test
- ANOVA
- Post Hoc analysis - Tukey's HSD
_____________________________________________________
25C fermentations ANOVA and Post Hoc analysis:

Levene's test
```{r message=FALSE, warning=FALSE}
filtered_df <- hot

#this does the test on each column of filtered_df starting at column 4
levene.results <- apply(filtered_df[,6:ncol(filtered_df)],2,function(x) {leveneTest(x~filtered_df$strain)})

#this makes a table of pvalues
levene.plvals<-lapply(levene.results,function(x) {c(x$`Pr(>F)`)}) %>% as.data.frame()
head(levene.plvals)
#Make an excel file for all the hot temp anova related results. 
#write.xlsx(levene.plvals, file="output/Spreadsheets/25C_output_res.xlsx", sheetName="LevenesTest", row.names=TRUE)
```

ANOVA
```{r}
#define df I am analyzing 
x<-hot

#now create an empty df
anova.out <- data.frame(ss_trt = rep(NA, 33), ss_error = rep(NA, 33), f_val = rep(NA, 33), p_val = rep(NA, 33), stringsAsFactors = FALSE)

row.names(anova.out) <- colnames(x[,5:ncol(x)])

#Create the for loop (compound ~ yeast strain) within in a single trt
for (i in 6:ncol(x)) {
lm.res<-lm(x[[i]]~ x$strain)
anova.res<-anova(lm.res)
anova.out$ss_trt[i-5] <- anova.res$`Sum Sq`[2]
anova.out$ss_error[i-5] <- anova.res$`Sum Sq`[1]
anova.out$f_val[i-5] <- anova.res$`F value`[1]
anova.out$p_val[i-5] <- anova.res$`Pr(>F)`[1]
}
anova.out  

#Write the results to an excel file inside a folder for this analysis. 
#write.xlsx(anova.out, file="output/Spreadsheets/25C_output_res.xlsx", sheetName="ANOVA_results", append = TRUE, row.names=TRUE)
```

For the post hoc analysis I will use a Tukey's HSD test and generate letter codes to indicate significance. 
Strains that share the same letter are not significantly different. 
```{r}
#Making a loop that will output the results into a df. 
x <- hot
#Make and empty df to fill
hsd_df.out <-data.frame(X2_phenylethanol = rep(NA, 11),
                        benzalcohol = rep(NA, 11), 
                        dihydroxy_hexamethoxyflavone_putative = rep(NA, 11), 
                        hexanoic_acid = rep(NA, 11),
                        octanoic_acid = rep(NA, 11),
                        X2_phenylethyl_acetate = rep(NA, 11),
                        methionol = rep(NA, 11),
                        X3_methylbutanoic_acid = rep(NA, 11),
                        X2.methylbutanoic_acid = rep(NA, 11),
                        ethyl_decanoate = rep(NA, 11),
                        acetic_acid_putative = rep(NA, 11),
                        ethyl_octanoate = rep(NA, 11),
                        hexanol = rep(NA, 11), 
                        hexyl_acetate = rep(NA, 11),
                        ethyl_hexanoate = rep(NA, 11),
                        X3methyl_butanol = rep(NA, 11),
                        X2methyl_butanol = rep(NA, 11),
                        X2methylbutyl_acetate = rep(NA, 11),
                        X2methylpropanol = rep(NA, 11),
                        ethyl_3_methyl_butanoate = rep(NA, 11),
                        ethyl_2_methyl_butanoate = rep(NA, 11),
                        ethyl_butanoate = rep(NA, 11),
                        ethyl_2_methyl_propanoate = rep(NA, 11),
                        ethyl_propanoate = rep(NA, 11),
                        ethyl_acetate = rep(NA, 11),
                        ethyl_formate_putative = rep(NA, 11),
                        acetaldehyde_putative = rep(NA, 11),
                        ethananolamine_putative = rep(NA, 11),
                        ethanol= rep(NA, 11),
                        glycerol= rep(NA, 11),
                        glucose= rep(NA, 11),
                        fructose= rep(NA, 11),
                        aucl = rep(NA, 11),
                        stringsAsFactors = FALSE) 
row.names(hsd_df.out) <- sort(unique(x$strain))

for (i in 6:ncol(x)) {
model <- aov(x[[i]]~strain, data = x)  
hsd_res <- HSD.test(model, "strain", group = TRUE)
groups <- data.frame(hsd_res$groups)
groups <- groups[order(row.names(groups)),]
hsd_df.out[,i-5] <- groups$groups
}
hsd_df.out
#Save the output into the anova results file:
#write.xlsx(hsd_df.out, file="output/Spreadsheets/25C_output_res.xlsx", sheetName="HSD_PostHoc_results", append = TRUE, row.names=TRUE)
```

_____________________________________________________
15C fermentation ANOVA and post hoc analysis:

Levene's test
```{r message=FALSE, warning=FALSE}
filtered_df <- cold

#this does the test on each column of filtered_df starting at column 4
levene.results <- apply(filtered_df[,6:ncol(filtered_df)],2,function(x) {leveneTest(x~filtered_df$strain)})

#this makes a table of pvalues
levene.plvals<-lapply(levene.results,function(x) {c(x$`Pr(>F)`)}) %>% as.data.frame()
levene.plvals
 
#Write an excel file that will keep all of our anova results from cold fermentation analysis:
#write.xlsx(levene.plvals, file="output/Spreadsheets/15C_output_res.xlsx", sheetName="LevenesTest", row.names=TRUE)
```


ANOVA for 15C (cold) fermentations
```{r}
#define df I am analyzing as x so that I dont have to change all the code for cold fermentaitons
x<-cold
#now create an empty df
anova.out <- data.frame(ss_trt = rep(NA, 33), ss_error = rep(NA, 33), f_val = rep(NA, 33), p_val = rep(NA, 33), stringsAsFactors = FALSE)
row.names(anova.out) <- colnames(x[,5:ncol(x)])
#Create the for loop (compound ~ yeast strain) within in a single trt
for (i in 6:ncol(x)) {
lm.res<-lm(x[[i]]~ x$strain)
anova.res<-anova(lm.res)
anova.out$ss_trt[i-5] <- anova.res$`Sum Sq`[2]
anova.out$ss_error[i-5] <- anova.res$`Sum Sq`[1]
anova.out$f_val[i-5] <- anova.res$`F value`[1]
anova.out$p_val[i-5] <- anova.res$`Pr(>F)`[1]
}
anova.out  
#Write the results to an excel file inside a folder for this analysis. 
#write.xlsx(anova.out, file="output/Spreadsheets/15C_output_res.xlsx", sheetName="ANOVA_results", append = TRUE, row.names=TRUE)
```

For the post hoc analysis I will use a Tukey's HSD test and generate letter codes to indicate significance. 
Strains that share the same letter are not significantly different. 
```{r}
#Making a loop that will output the results into a df. 
x <- cold
#Make and empty df to fill
hsd_df.out <-data.frame(X2_phenylethanol = rep(NA, 11),
                        benzalcohol = rep(NA, 11), 
                        dihydroxy_hexamethoxyflavone_putative = rep(NA, 11), 
                        hexanoic_acid = rep(NA, 11),
                        octanoic_acid = rep(NA, 11),
                        X2_phenylethyl_acetate = rep(NA, 11),
                        methionol = rep(NA, 11),
                        X3_methylbutanoic_acid = rep(NA, 11),
                        X2.methylbutanoic_acid = rep(NA, 11),
                        ethyl_decanoate = rep(NA, 11),
                        acetic_acid_putative = rep(NA, 11),
                        ethyl_octanoate = rep(NA, 11),
                        hexanol = rep(NA, 11), 
                        hexyl_acetate = rep(NA, 11),
                        ethyl_hexanoate = rep(NA, 11),
                        X3methyl_butanol = rep(NA, 11),
                        X2methyl_butanol = rep(NA, 11),
                        X2methylbutyl_acetate = rep(NA, 11),
                        X2methylpropanol = rep(NA, 11),
                        ethyl_3_methyl_butanoate = rep(NA, 11),
                        ethyl_2_methyl_butanoate = rep(NA, 11),
                        ethyl_butanoate = rep(NA, 11),
                        ethyl_2_methyl_propanoate = rep(NA, 11),
                        ethyl_propanoate = rep(NA, 11),
                        ethyl_acetate = rep(NA, 11),
                        ethyl_formate_putative = rep(NA, 11),
                        acetaldehyde_putative = rep(NA, 11),
                        ethananolamine_putative = rep(NA, 11),
                        ethanol= rep(NA, 11),
                        glycerol= rep(NA, 11),
                        glucose= rep(NA, 11),
                        fructose= rep(NA, 11),
                        aucl = rep(NA, 11),
                        stringsAsFactors = FALSE) 
row.names(hsd_df.out) <- sort(unique(x$strain))

for (i in 6:ncol(x)) {
model <- aov(x[[i]]~strain, data = x)  
hsd_res <- HSD.test(model, "strain", group = TRUE)
groups <- data.frame(hsd_res$groups)
groups <- groups[order(row.names(groups)),]
hsd_df.out[,i-5] <- groups$groups
}
hsd_df.out
#write output to the excel file
#write.xlsx(hsd_df.out, file="output/Spreadsheets/15C_output_res.xlsx", sheetName="HSD_PostHoc_results", append = TRUE, row.names=TRUE)
```
___________________________________________________________________________________________
___________________________________________________________________________________________
Heatmaps

```{r}
library(RColorBrewer)
#library(ComplexHeatmap)
```

```{r}
suppressWarnings({
#15C all compounds averaged 
cold_averages<-read_xlsx("15C.xlsx", sheet= "averages_for_correlation") %>% data.frame()
rnames<- cold_averages$strain
cold_averages <- cold_averages[2:ncol(cold_averages)]
row.names(cold_averages) <- rnames
cold_averages <- as.matrix(cold_averages)

#25C all compounds averaged
hot_averages<-read_xlsx("25C.xlsx", sheet= "averages_for_correlation") %>% data.frame()
rnames<- hot_averages$strain
hot_averages <- hot_averages[2:ncol(hot_averages)]
row.names(hot_averages) <- rnames
hot_averages <- as.matrix(hot_averages)

#Sclaed by column,since we need to absorb the variation between column.
trans<- t(as.matrix(hot_averages))

trans<- t(as.matrix(cold_averages))

x <- trans
rm <- rowMeans(x)
x             <- sweep(x, 1, rm)
sx            <- apply(x, 1, sd)
x             <- sweep(x, 1, sx, "/")


heatmap(x, name = "x", 
        col = brewer.pal(11,"RdYlBu"),
        cluster_rows = FALSE, cluster_columns = FALSE,
        row_split = rep(c("A", "B", "C", 
                          "D", "E" ), 
                        c(4,5,7,8,4)),
        cluster_row_slices = FALSE, cluster_column_slices = FALSE)
})

```



